Types for R, Empirically Designed and Evaluated

### 2020-05-07 paper/main.tex
Fix challenge description for contractr

%
%
\subsubsection{Challenges}
There are two interesting challenges that need to be tackled to make this work
reliably, without surprises for the end-users. Firstly, we need to correctly
type-check arguments. R is a non-strict language~\cite{oopsla19};
when a function is called, instead of getting values as arguments which can be
immediately type-checked, we get unevaluated code thunks. We can force the
evaluation of these thunks to obtain values for type-checking but that will not
preserve the language's semantics. This will lead to incorrect results and
surprises for the end-users. To preserve the original semantics, we modify the
thunks to call our type-checker on the value obtained when they are forced,
along with the context necessary to associate the thunk to the function and its
parameter (for failure messages). This works, but with one wrinkle. GNU R has a
bytecode compiler that can sometimes optimize away thunks, in those cases, we
don't get thunks as arguments but values that can be immediately type-checked.
However, we still can't type-check the value and issue failure messages because
that would violate the semantics. Non-strict semantics require that any
computation related to the argument happens on the first use of the argument. If
this argument was never used, it would be incorrect to type-check it. To make
this work correctly, we explicitly wrap the argument value in a thunk with a
call to the type-checker as in the previous case.  Secondly, we need to
correctly type-check function return values. In R, the result of function
evaluation is the result of evaluating the last expression in its body; so a
return call is absent from most function definitions. So there are many
potential subexpressions which need to be type-checked aginst the return type.
For this, we register our type-checker to be called on the return value in a
function exit hook. This hook is executed in the function's environment after it
has finished executing. While this works, we also have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes R functions on the stack to be discarded. When they are discarded,
their exit hooks are called, which in our case calls the type-checker. But,
these functions are in the middle of an active computation, so they don't have a
return value to type-check. We deal with this problem by allocating a unique
object which serves as the return value for calls that are discarded. Our exit
hook does not call the type-checker if this unique value happens to be the
return value of the function.

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to contractr in three ways: as part of
its internal database of types (which already has type signatures for our corpus
of 500 packages), as part of a designated TYPEDECLARATION file supplied by the
package author and installed along with the package, or through a user-level API
function insert\_contract that lets the user insert contract to a custom
function with the type declaration provided as a string argument.

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design.
Firstly, the package enables type-checking for all packages with type
signatures, automatically, on importing, offering a seamless and hassle-free
experience for both package developers and end-users. A mere library(contractr)
is enough to insert contracts, both, in packages pre-loaded in the user's
workspace and in packages that will be loaded eventually. For this, contractr
relies heavily on R's reflective and dynamic capabilities. Upon loading
contractr scans all the pre-loaded packages in the user's workspace and inserts
contracts in functions for which type signatures are available. For all other
packages installed on user's machine but not yet loaded in the workspace,
contractr sets up package load hooks. The hooks get executed when these packages
get loaded and insert contracts to the package functions. The ensures that users
do not fiddle manually with the contractr API to enable type-checking for
packages loaded eventually. Furthermore, contractr automatically removes
contracts from all the package functions and restores them to their original
state when it is unloaded, an operation that is rarely performed but is
possible, nonetheless.
Secondly, we enable package authors to supply type declarations to contractr for
their package functions, without requiring any modification to package code.
This is accomplished by providing type declarations inside roxygen2 function
comment blocks. roxygen2 is an R package that enables authors to add
documentation to R functions in the form of organized plaintext sections which
are automatically exported to R's latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. contractr uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up contractr when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type declaration:

This technique enables the function and its type signature to coexist next to
each other, where they are more likely to remain synchronized. It also enables
package authors to use our type system without modifying their code.
Thirdly, we provide a very expressive API and package switches to the users,
covering a variety of uses cases. During development, developers can explicitly
insert contracts by supplying the type declaration as a string argument to the
insert\_contract API function. Contracts can be removed by calling
remove\_contract on a function. Contracts can be selectively enabled or disabled
for code blocks by wrapping them in calls to ignore\_contracts and
capture\_contracts functions. This enables developers and users to selectively
type check sections of code. Furthermore, these functions return a data frame
that contains all the information about failed and successful contract
assertions in the wrapped code blocks. Multiple type-checking failures can
introduce a lot of noise in the program output. To alleviate this problem, users
can set a flag, contractr.severity, to “silence” which suppresses the failure
messages. The contract assertions can then be obtained as a data frame through a
call to get\_contracts at any program point. Setting this flag to “error” will
turn the type-checking failure warnings to errors.

>>>>

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to \emph{contractr} in three ways: as
part of its internal database of types (which already has type signatures for
our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
supplied by the package author and installed along with the package, or through
a user-level API function insert\_contract that lets the user insert contract to
a custom function with the type declaration provided as a string argument.

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \emph{contractr}.
Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. When a function is called in
R, parameters are bound to unevaluated code thunks called
promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value for
type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \emph{contractr} modifies the
promise by wrapping the unevaluated promise expression in a call to its
type-checker along with the context necessary to associate the promise to the
function and its parameter (for failure messages). The contract checking happens
when the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \emph{contractr} receives
values that can be immediately type-checked. Non-strict semantics dictate that
any computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would be
incorrect to type-check it at the beginning of call. To make this work,
\emph{contractr} mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.
Secondly, we faced the problem of type-checking function return value. In R, the
result of a function call is the result of evaluating the last expression in the
function body; so an explicit return call is absent from most function
definitions. Thus, there could be many potential sub-expressions in the function
body which need to be type-checked aginst the return type. To address this,
\emph{contractr} registers the type-checker to be called on the return value
through a function exit hook. This hook is executed in the function call's
environment after it has executed. While this works, we have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes active R function calls on the stack to be discarded. When they are
discarded, their exit hooks are called, which in this case calls the registered
type-checker. But, these functions are in the middle of an active computation,
so they don't have a return value to type-check. \emph{contractr} deals with
this problem by allocating a unique sentinel object which serves as the return
value for calls that are discarded. The function call exit hook does not call
the type-checker if this unique value happens to be the return value of the
call.

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design. Firstly, the package enables type-checking for all
packages with type signatures, automatically, on importing, offering a seamless
and hassle-free experience for both package developers and end-users. A mere
library(\emph{contractr}) is enough to insert contracts, both, in packages
pre-loaded in the user's workspace and in packages that will be loaded
eventually. For this, \emph{contractr} relies heavily on R's reflective and
dynamic capabilities. Upon loading \emph{contractr} scans all the pre-loaded
packages in the user's workspace and inserts contracts in functions for which
type signatures are available. For all other packages installed on user's
machine but not yet loaded in the workspace, contractr sets up package load
hooks. The hooks get executed when these packages get loaded and insert
contracts to the package functions. The ensures that users do not fiddle
manually with the \emph{contractr} API to enable type-checking for packages
loaded eventually. Furthermore, \emph{contractr} automatically removes contracts
from all the package functions and restores them to their original state when it
is unloaded, an operation that is rarely performed but is possible, nonetheless.
Secondly, we enable package authors to supply type declarations to
\emph{contractr} for their package functions, without requiring any modification
to package code. This is accomplished by providing type declarations inside
roxygen2 function comment blocks. roxygen2 is an R package that enables authors
to add documentation to R functions in the form of organized plaintext sections
which are automatically exported to R's latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. \emph{contractr} uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up \emph{contractr} when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type
declaration:

This technique enables the function and its type signature to coexist next to
each other, where they are more likely to remain synchronized. It also enables
package authors to use our type system without modifying their code. Thirdly, we
provide a very expressive API and package switches to the users, covering a
variety of uses cases. During development, developers can explicitly insert
contracts by supplying the type declaration as a string argument to the
insert\_contract API function. Contracts can be removed by calling
remove\_contract on a function. Contracts can be selectively enabled or disabled
for code blocks by wrapping them in calls to ignore\_contracts and
capture\_contracts functions. This enables developers and users to selectively
type check sections of code. Furthermore, these functions return a data frame
that contains all the information about failed and successful contract
assertions in the wrapped code blocks. Multiple type-checking failures can
introduce a lot of noise in the program output. To alleviate this problem, users
can set a flag, \emph{contractr}.severity, to “silence” which suppresses the
failure messages. The contract assertions can then be obtained as a data frame
through a call to get\_contracts at any program point. Setting this flag to
“error” will turn the type-checking failure warnings to errors.
